Refining from the VHDL-AMS Hardware Description Language to the BON High-level Specification Language

### 2010-01-04 papers/BON_VHDL/paper.tex
New abstract written in OmmWriter.  Revised abstract and most of introduction.

The \textbf{UCD CASL SenseTile System} is a large-scale,
general-purpose sensor system installed at the University College
Dublin in Dublin, Ireland.  \textbf{Our facility provides a
capability---unique in the world in terms of its scale and
flexibility---for performing in-depth investigation of both the
specific and general issues in large-scale sensor networking.}

This system integrates a sensor platform, a datastore, and a compute
farm.  The sensor platform is a custom-designed, but inexpensive,
sensor platform (called the \ST) paired with general-purpose
small-scale compute nodes, including everything from low-power PDAs to
powerful touchscreen-based portable computers.  The datastore is a
multi-terabyte scale scientific datastore into which sensor data
flows, and in which online and offline scientific computation of
sensor and other scientific data takes place.  The compute farm is a
large-scale Linux, Solaris, and OS~X-based compute farm used by
scientists for these online and offline scientific computations.

The \STs, as an embedded system, have several unusual built-in sensors
and emitters, sensor nodes can be extended via a general-purpose USB
bus, and \STs contain an FPGA that is dynamically updated on a
per-experiment basis.  \STs are custom-designed in VHDL-AMS.  Like
many hardware-based systems, to write new software for or against the
\ST architecture one either uses the actual hardware or one runs a
\STS.  But this simulator emulates a piece of hardware, one would like
to know if the simulator \emph{actually} simulates the hardware.  We
have developed an approach through which we refine the VHDL-AMS
specification up to the formal BON specification language, and from
this specification we generate an executable, functional model-based
specification in the JML language.  We have used this specification to
implement a fully validated and verified \ST simulator.

@@

>>>>

Complex software engineering projects often use high-level
specification languages to describe system abstractions.  These
abstractions, which typically capture static and dynamic architectural
aspects, are meant to be more approachable, comprehensible, and useful
than witnessing the gory details of the systems they describe.  In the
hardware world, systems are described with hardware description
languages like VHDL.  These languages exist at a similar level of
abstraction as traditional software programming languages, like Java.
But whence are the high-level hardware description languages?
Moreover, if one is doing hardware/software co-design, what
architecture description languages are usable and useful to both
facets of the problem domain?  In this paper, we describe a
non-trivial co-design effort, called SenseTile, and how requirements
on its design and verification led us to define a formal refinement of
VHDL-AMS to the EBON high-level specification language.  This
refinement enables us to abstract away, in a seamless, reversible, and
contract-based fashion, from the details of a particular hardware
system to an neutral architecture description which, in turn, can be
refined to an executable Java-based software simulator.  This pair of
refinements permits full round-trip hardware/software co-design with
verification, as the Java-based system is annotated with formal
specifications using the Java Modeling Language.

@@

The \textbf{UCD CASL SenseTile System} is a large-scale,
general-purpose sensor system under development at the University
College Dublin in Dublin, Ireland.  \emph{After its installation, it
  is our intention that this facility will provide a
  capability---unique in the world in terms of its scale and
  flexibility---for performing in-depth investigation of both the
  specific and general issues in large-scale sensor networking.}  This
system integrates a \emph{sensor platform}, a \emph{datastore}, and a
\emph{compute farm}.

The \emph{sensor platform} is a custom-designed, but inexpensive,
sensor platform called the \ST.  Each \ST is paired with
general-purpose, small-scale compute node ranging is capability from a
low-power PDA to a powerful touchscreen-based portable computer.

The \emph{datastore} is a multi-terabyte scale scientific datastore
located in a server room across campus.  Sensor data flows into this
datastore during experiments.  It is also used for online and offline
computation of sensor and other scientific data.

The \emph{compute farm} is a large-scale Linux, Solaris, Windows, and
OS~X-based compute farm used by scientists for these computations.

\subsection{The \ST}
\label{subsec:st}

The \STs contain several built-in sensors, ranging from typical
sensors like temperature and humidity to more unusual ones like
ultrasonic microphones.  They also contain several emitters, including
visible and infrared light, audible and ultrasonic sound, and more.
Sensor nodes are extended via a USB bus, and \STs contain an FPGA that
is dynamically updated on a per-experiment basis.

\STs are custom-designed by a third party based upon our high-level
specifications in VHDL-AMS.  Common to many hardware prototype
development efforts, writing software for the \ST architecture means
either using the actual hardware, or running a \STS.  But this
simulator must emulate a piece of hardware \emph{that does not yet
  exist}.  Worse, as is typical with these kinds of development
efforts, the hardware is several quarters late, and the show must go
on.

Consequently, the \STS becomes the primary focus of development
efforts and, as we do not wish to loose any development time, we would
like to know if the simulator \emph{actually} simulates the hardware,
thus ensuring that the software we write for the simulator will run
immediately on the hardware when it arrives.

\subsection{Software}
\label{subsec:software}

Complementing these hardware development efforts, the software
architecture that runs the \STs is under development.  This system
consists of a low-level USB driver written in C and a packet framer
and control system written in Java.

It is vitally important that these low-level software systems are
well-understood, work correctly from day zero, and have adequate
non-functional characteristics---primarily that they do not use too
much memory or CPU on the low-power devices for which they are
destined.  As such, and following the standard development conventions
of our research group, they are formally specified using the BON and
JML specification languages.  Moreover, they are validated and
verified using various forms of testing and static analysis, all of
which are the focus of several other papers.

\subsection{Co-design}
\label{subsec:co-design}

Even in the best of hardware-software co-design processes, hardware
development typically focuses on correctness through rigorous
specification, static analysis, and simulation, while software
development uses the ``by the seat of the pants'' approach.  As is
obvious from the previous section, that does not fly in our research
group.

Consequently, we have developed an approach through which we abstract
(via a formal refinement) a VHDL-AMS specification up to a formal BON
specification.  Using this specification, we then generate (again, via
a formally defined refinement) an executable, functional, model-based
specification in the JML language.  We have subsequently used this
specification to implement a fully validated and verified \ST
simulator.  Due to the semantic properties of these refinements, as
seen in~\autoref{diagram}, we are guaranteed that all functional
properties of the original VHDL-AMS specification are preserved in the
software simulator, thereby witnessing full hardware-software
co-design in a formal sense.

XXX
